---
title: "Dirty Data Roundtable"
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyverse)
library(readxl)
```


## Task 1

Task 1
- Jack, Lloyd, Chris


Decathlon Events
a bunch of athletes and their scores from values from two competitions

### Cleaning

not any missing values


a bit of wrangling required for analysis

pretty much the same

cleaning steps

1. read
2. clean up column names
3. convert row names into column of athelte names
4. convert athlete names to upper case
5. (optional) pivot to longer form
6. write to file

```{r}
decathlon_raw <- read_rds("raw_data/decathlon.rds")
```


```{r}
decath_clean <- decathlon_raw %>% 
  janitor::clean_names() %>% 
  rownames_to_column("athlete") %>% 
  mutate(athlete = str_to_upper(athlete)) %>% 
  # pivot_longer(-c(athlete, rank, points, competition), names_to = "event", values_to = "time_or_distance") %>% 
  write_csv("clean_data/decathlon_results.csv")
  # rename_with(.cols = starts_with("x"), .fn = ~str_remove(.x, "x")) %>% 
```


### Analysis

3. Who had the highest total points across both competitions?

```{r}
decath_clean %>% 
  group_by(athlete) %>% 
  summarise(total_points = sum(points)) %>% 
  slice_max(total_points)
```

If the data had been pivoted: 

```{r}
decath_clean %>% 
  pivot_longer(-c(athlete, rank, points, competition), names_to = "event", values_to = "time_or_distance") %>% 
  group_by(athlete, competition) %>% 
  summarise(points = min(points), .groups = 'drop') %>% 
  slice_max(points, n = 1)
  
```


## Task 2
- Sarah, Jamie

about cakes

2 tables (1 with ingredient info, 1 with recipes)

### Cleaning

data needs to be combined
questions all concerned ingredients

1. converted to long form (ingredient code to a column, amounts to another one)
2. join with cake codes
3. dealt with missing values

```{r}
cake_ingredients <-
  read_csv("raw_data/cake/cake-ingredients-1961.csv")

cake_ingredients_codes <-
  read_csv("raw_data/cake/cake_ingredient_code.csv")
```


```{r}
# missing measure for sour cream cup 
cake_ingredients_codes_clean <- cake_ingredients_codes %>% 
  mutate(
    measure = if_else(
      ingredient == "Sour cream cup", "cup", measure
      ),
    ingredient = if_else(
      ingredient == "Sour cream cup", "Sour cream", ingredient
      ))

cake_ingredients %>% 
  pivot_longer(-Cake, names_to = "ingredient_code", values_to = "amount", values_drop_na = TRUE)  %>% 
  left_join(cake_ingredients_codes_clean, by = c("ingredient_code" = "code")) -> cake_ingredients_clean
  # assumption that NA ingredient values mean that ingredient is not in the recipe for the cake 
  

```




### Analysis

4. Which cake has the most unique ingredients?

which cake has the most ingredients that aren't in any other cake

which cake has the most ingredients that aren't the same (the most types of ingredients)


```{r}
# David's approach for Jamie question
cake_ingredients_clean %>% 
  # 1. get the number of times each ingredient was used
  group_by(ingredient_code) %>% 
  mutate(n_uses = n()) %>% 
  ungroup() %>% 
  filter(n_uses == 1) %>% 
  group_by(Cake) %>% 
  summarise(n_unique_y_ingredients = n()) %>% 
  slice_max(n_unique_y_ingredients)
```

```{r}
cake_ingredients_clean %>% 
  group_by(Cake) %>% 
  summarise(n_ingredients = n_distinct(ingredient)) %>% 
  slice_max(n_ingredients)
``` 


```{r}
ingreds <- c("flour", "flour", "eggs", "baking_soda")

length(ingreds)
n_distinct(ingreds)
```




## Task 3
- Nico, Rob, Neil


two tables relating to bird sightings

1. ship location
2. bird sightings
(quite large)

some data was more relevant for the analysis

### Cleaning

```{r}
library(readxl)

excel_sheets("raw_data/seabirds.xls")

ship_data <- read_xls("raw_data/seabirds.xls", 1)
bird_data <- read_xls("raw_data/seabirds.xls", 2, guess_max = 30000)

# 1. read in
# 2. clean column names
# 3. clean bird data
# 3.1 dropped extra columns
# 3.2 trim column names
# 3.3 wrangled time column

# 4. clean ship data

# 5. join tables
```

Data dictionary

```{r}
read_xls("raw_data/seabirds.xls")
```



```{r}
# species common name, scientific name, latitude, count, joining columns

library(lubridate)

ship_subset <- ship_data %>% 
  janitor::clean_names() %>% 
  select(record, record_id, date, time, lat)


birds_clean <- bird_data %>% 
  janitor::clean_names() %>% 
  select(record, record_id, 3, 4, 5, age:count) %>% 
  left_join(ship_subset, by = c("record_id"), suffix = c("_bird", "_ship")) %>% 
  mutate(time = str_remove(time, "^\\d{4}-\\d{2}-\\d{2} ")) %>% 
  rename(
    species_common_name = species_common_name_taxon_age_sex_plumage_phase,
    species_scientific_name = 
      species_scientific_name_taxon_age_sex_plumage_phase
  )
```

## Analysis

How many different types of birds were only ever seen in groups of 1?

```{r}
birds_clean %>% 
  mutate(species_common_name = str_remove_all(species_common_name, "[A-Z0-9 ]{2,}")) %>% 
  mutate(species_common_name = str_remove(species_common_name, "sensu lato *^")) %>% 
  group_by(species_common_name) %>%
  mutate(max_observed_group = max(count, na.rm = TRUE)) %>% 
  filter(max_observed_group == 1) %>% 
  distinct(species_common_name)
```




## Task 4

- Everyone


dataset involving trick or treat survey data from 3 halloweens
(2015, 2016, and 2017)

as part of the survey people rated certain candies

```{r}
# each dataset starts with > 1000 rows and > 100 columns

candy_data_2015 <- read_xlsx("raw_data/candy_ranking_data/boing-boing-candy-2015.xlsx") %>% 
  janitor::clean_names()

candy_data_2016 <- read_xlsx("raw_data/candy_ranking_data/boing-boing-candy-2016.xlsx") %>% 
  janitor::clean_names()

candy_data_2017 <- read_xlsx("raw_data/candy_ranking_data/boing-boing-candy-2017.xlsx") %>% 
  janitor::clean_names()
```


### Cleaning

0. with initial exploration
1. remove unnecessary columns (after some data exploration)
2. pivot to long format (clean up 2017 columns)
* cleaned up the age column
3. combined the three datasets

+ 1

box_o_raisins
boxo_raisins

Goal format

| year | age | country | trick_treating| gender| candy | rating |

```{r}
# 2015

# things to come back to maybe (other candy that wasn't in the
# survey that brought joy)

clean_2015 <- candy_data_2015 %>% 
  mutate(id = str_c(2015, row_number(), sep = "_")) %>% 
  pivot_longer(butterfinger:york_peppermint_patties, names_to = "candy", values_to = "rating") %>% 
  select(
    id,
    age = how_old_are_you,
    trick_treating = are_you_going_actually_going_trick_or_treating_yourself,
    candy,
    rating
  ) %>% 
    mutate(year = 2015)
```


```{r}
# 2017

clean_candy_2017 <- candy_data_2017 %>% 
  mutate(id = str_c(2017, row_number(), sep = "_")) %>% 
  pivot_longer(starts_with("q6"), names_to = "candy", values_to = "rating", names_prefix = "q6_") %>% 
  rename_with(.fn = ~str_remove(.x, "q[0-9]+_"),
              .cols = everything()
  ) %>% 
  select(id, trick_treating = going_out, gender, age, country, candy, rating) %>% 
  mutate(year = 2017)
```


```{r}
clean_candy_2016 <- candy_data_2016 %>% 
  mutate(id = str_c(2016, row_number(), sep = "_")) %>% 
  pivot_longer(x100_grand_bar:york_peppermint_patties,
              names_to = "candy", values_to = "rating") %>% 
  select(
    id,
    trick_treating = are_you_going_actually_going_trick_or_treating_yourself,
    gender = your_gender,
    age = how_old_are_you,
    country = which_country_do_you_live_in,
    candy,
    rating
  ) %>% 
  mutate(year = 2016)
```


Task 1

explore the dataset
clean column names
check for missing values
determine what variables will be required for analysis

pivoting and or joining to be in a suitable form for analysis


```{r}
# Jamie's treatment of NA ratings / age + trick_treating
clean_candy_2017 <- clean_candy_2017 %>% 
  filter(!is.na(rating) | (!is.na(age) & !is.na(trick_treating)))

clean_candy_2016 <- clean_candy_2016 %>% 
  filter(!is.na(rating) | (!is.na(age) & !is.na(trick_treating)))

clean_candy_2015 <- clean_2015 %>% 
  filter(!is.na(rating) | (!is.na(age) & !is.na(trick_treating)))
```


- joining the datasets together into one: clean_candy_long

```{r}
all_candy <- bind_rows(
  list(clean_candy_2015, clean_candy_2016, clean_candy_2017)
)
```

cleaning up the countries

- constantly referring to

```{r}
# country counts
all_candy %>% 
  distinct(id, .keep_all = TRUE) %>% 
  count(country)
```


```{r}
clean_country <- all_candy %>% 
  mutate(country = str_to_lower(country)) %>% 
  mutate(country = str_remove_all(country, "[:punct:]")) %>% 
  mutate(country = case_when(
    str_detect(country, "^us|united state|unites states|[ue]rica|units states") ~ "usa",
    str_detect(country, "^united k|uk|en[dg]land|scotland|wales") ~ "uk",
    str_detect(country, "^can") ~ "canada",
    is.na(country) ~ NA_character_,
    TRUE ~ "other"
  ))
```


cleaning up the countries

- constantly referring to

```{r}
# country counts
all_candy %>% 
  count(candy)
```

Thoughts: 

100 grand bar (check for any with numbers at the start and an x)
box_o_raisins or boxo_raisins

```{r}
clean_candy <- clean_country %>% 
  mutate(candy = case_when(
    str_detect(candy, "100_grand") ~ "100_grand_bar",
    str_detect(candy, "box[_]*o_raisins") ~ "box_o_raisins",
    str_detect(candy, "^anonymous_brown_globs") ~ "anonymous_brown_globs",
    TRUE ~ candy
  ))
```



so far we've 

-- > reduce data size [x]
-- > combine into one file [x]
-- > country column [x]
-- > candy column [x]

still to do 

-- > age column [] (probably would do this before cleaning the country column - since there is a little bit of data to grab from there)

if we just do `as.numeric(age)`:

"25-ish" -- > NA
"25.5" -- > 25.5

```{r}

clean_ages <- clean_candy %>% 
  mutate(age_extract = str_extract(age, "\\d+\\.*\\d*")) %>% 
  mutate(age = as.numeric(age)) %>% 
  filter(age > 0, age < 110)
```


### Analysis

if you didn't pivot, a lot of your analysis will look like

```{r}
clean_data %>% 
  summarise(across(candy_columns, your_function))
```


What is the total number of candy ratings given across the three years. (Number of candy ratings, not the number of raters. Don’t count missing values)

```{r}
clean_ages %>%
  drop_na(rating) %>% 
  nrow()
```


What was the average age of people who are going out trick or treating?


```{r}
clean_ages %>% 
  filter(trick_treating == 'Yes') %>%
  distinct(id, .keep_all = TRUE) %>% 
  summarise(avg_age = round(mean(age, na.rm= TRUE)))
```


```{r}
unique(clean_ages$age) #checking change to numeric coerced outliers to NA
unique(full_data$going_out) 

clean_ages %>% 
  filter(trick_treating == "Yes", 
         age != "Inf",
         age <= 118) %>% #oldest living person
  summarise(avg_age = mean(age)) %>% 
  round()
```

For each of joy, despair and meh, which candy bar received the most of these ratings?


```{r}
# single pipe
clean_ages %>% 
  group_by(rating, candy) %>% 
  summarise(n_rating_per_candies = n()) %>% 
  slice_max(n_rating_per_candies, n = 1)

# multiple table solution
clean_ages %>% 
  filter(rating == "DESPAIR") %>% 
  group_by(candy, rating) %>% 
  summarise(joy_rating_per_candies = n()) %>% 
  arrange(desc(joy_rating_per_candies)) %>% 
  head(1) %>% 
  ungroup()

candy %>% 
  filter(rating == "MEH") %>% 
  group_by(candies, rating) %>% 
  summarise(joy_rating_per_candies = n()) %>% 
  arrange(desc(joy_rating_per_candies)) %>% 
  head(1) %>% 
  ungroup()
```


```{r}
clean_ages %>% 
  select(candy, rating) %>% 
  #groups by ratings and then type of candy
  group_by(rating, candy) %>% 
  #counts the number of ratings for each rating category for each candy
  summarise(num_ratings = n()) %>% 
  #takes the highest rated candy for each rating.
  slice_max(order_by = num_ratings, n = 1)
```



How many people rated Starburst as despair?


```{r}
clean_ages %>% 
  filter(candy == "starburst",
         rating == "DESPAIR") %>% 
  summarise(count = n())
```




## Priorities for the weekend

- rest and relaxation after a long project

- optional homeworks 
- PDA (week 2 assignment - meteors and unit testing, week 3 - sql meteors) 

## Next week

Data Visualisation with `{ggplot2}` and interactive data dashboards

