
# Seabirds Project Neil Plenderleith CodeClan DR16

## 1. Brief Introduction to the Dataset

The dataset is in .xlx format with 4 sheets. The main two sheets are bird sightings and the ship they are sighted from. There are also 2 sheets which describe the column data and abbreviations etc. There are a lot of NA values! I can understand a lot of these will be from incomplete sightings where not all data points can possible be collected from every sighting.

## 2. List of Assumptions

1. I did initially trim 3 very nearly empty columns thinking that I would leave the rest of the columns. However, after getting to know the data a bit better and completing the tasks I ended up removing much of the data - should it be necessary in the future this could be joined back in. 

2. Likewise I haven't gone crazy with NA values as I believe these tell a story in this data and the nature of not being able to collect all data points for every sighting.

3. I have assumed that the very large numbers in the bird_count column are ok. Having googled
the birds it appear they can be in colonies of 3 million so an estimated sighting of 99999 is perhaps not as crazy as it looked at first sight. Similarly the values 10000 - 99999 have been left out as well as these numbers of sightings must exist in colonies of such numbers.

## 3. Steps to Clean Data

### Load the required libraries, read in the data and join the two tables
```{r}
# Load Libraries ---------------------------------------------------------------

library(readxl)
library(here)
library(tidyverse)
library(janitor)

# Load Data ---------------------------------------------------------------

birds <- read_xls(here("raw_data/seabirds.xls"), 
                  sheet = "Bird data by record ID")
ships <- read_xls(here("raw_data/seabirds.xls"), 
                  sheet = "Ship data by record ID")

# Join Tables -------------------------------------------------------------

seabirds <- left_join(birds, ships, "RECORD ID")

```

### Next I got a look at the data and made a plan for cleaning. 
```{r}
# Investigate Data --------------------------------------------------------

glimpse(seabirds)
summary(seabirds)
skimr::skim(seabirds)

na_count <- seabirds %>%
summarise(across(.fns = ~ sum(is.na(.x)))) 
#this counts N/A values across all columns
```

### Cleaning the data
1. Sort column names
2. Drop 3 columns with very few values out of 49019 (4 more cols were borderline but we will keep these for now)
3. Change names of 2 long column names
4. Rename the count function to bird_count - too easily confused with the count function
5. Sort time column - it currently displays and old date as well as time
6. UPDATE - in the end I dropped a few columns from the data that were not specifically required by the tasks - these could always be added on later should the need arise
7. Investigate bird_count column - there are some high values in there!
```{r}

# Cleaning Data -----------------------------------------------------------

# Sort all these uppercase column names
seabirds <- clean_names(seabirds)

# Drop cols sex, sal and plphase very few values
seabirds <- seabirds %>% 
  select(-sex, -sal, -plphase)

# Change some long column names
seabirds <- seabirds %>% 
rename(species_common_name = species_common_name_taxon_age_sex_plumage_phase,
       species_scientific_name = species_scientific_name_taxon_age_sex_plumage_phase,
       bird_count = count) # count is easily confused with the function

# Lets tidy up the time column to just display the time and not the (incorrect) date
seabirds <- seabirds %>% 
  mutate(time = str_remove(time, "^[0-9]{4}-[0-9]+-[0-9]+ "))

# Lose some more non - essential columns
seabirds <- seabirds %>% 
  select(-(nfeed:record_y)) %>% 
  select(-(ew:depth)) %>% 
  select(-(csmeth:longecell), -wanplum)

# investigate bird_count column
seabirds %>% 
count(bird_count) %>% 
slice_max(bird_count, n = 20)

```
Looking at the bird_count column - there are some huge estimates for bird sightings. Looking at the codes in the raw data the bird_count column is birds spotted within a 10 minute window - is 99999 even possible? Its difficult here to decide what to keep in and what to take out? Doing some googling reveals the largest colony of shearwaters is 3 million so 99999 is perhaps understandable. Lets leave these big values as they stand.

Can we use this data now? It has a lot of NAs but i suppose a lot are just the way the data is recorded and the way the dataset is setup. Not every observation would have all fields completed.

### Finish by writing our cleaned data
```{r}
seabirds %>% 
write_csv(here("clean_data/seabirds.csv"))
```

## 4. Tasks


### Libraries and data

library(tidyverse)
```{r}
library(tidyverse)
library(readr)
library(here)

seabirds <- read_csv(here("clean_data/seabirds.csv"))

```

### Task 1 Which bird had the most individual sightings?

```{r}
# group by bird name and take a count of occurrences, join this back into the table
# to allow more names info to be pulled, arrange and slice head to get top hit
seabirds %>% 
  group_by(species_common_name) %>% 
  summarise(count_of_sightings = n()) %>%
  right_join(seabirds, "species_common_name") %>% 
  select(species_common_name, 
         species_scientific_name, 
         species_abbreviation, count_of_sightings) %>%
  arrange(desc(count_of_sightings))  %>% 
  slice_head()
# Answer:Wandering albatross sensu lato PL2, count 3104
```

### Task 2 Which bird had the highest total count?

```{r}
# Note i've taken this to mean highest count (adding the count of all sightings)
# And not the highest count at a single sighting

# group by bird name and take a sum of count, join this back into the table
# to allow more names info to be pulled, arrange and slice head to get top hit
seabirds %>% 
  drop_na(bird_count) %>% 
  group_by(species_common_name) %>% 
  summarise(total_count = sum(bird_count)) %>% 
  right_join(seabirds, "species_common_name") %>% 
  select(species_common_name, 
         species_scientific_name, 
         species_abbreviation, total_count) %>%
  arrange(desc(total_count)) %>% 
  slice_head(n=1)
  
# Answer: Short-tailed shearwater, count = 982553
```

### Task 3 Which bird had the highest total count above a latitude of -30?

```{r}
# same as Q2 but with a filter before for latitude > -30
seabirds %>% 
  filter(lat > -30) %>% 
  group_by(species_common_name) %>% 
  summarise(total_count = sum(bird_count)) %>%
  right_join(seabirds, "species_common_name") %>% 
  select(species_common_name, 
         species_scientific_name, 
         species_abbreviation, total_count) %>%
  arrange(desc(total_count)) %>% 
  slice_head()
# Answer: A Wedge-tailed shearwater, count 855
```

### Task 4 How many different types of birds were only ever seen in groups of 1?

```{r}
# ok so theres got to be a better way but i ended up adding two columns on
# with the individual sightings and the total count. If these were equal with no
# NA values which were dropped then all must have been sightings of one bird

# make our two table with the count and sum we need
seabirds_counting <- seabirds %>% 
  drop_na(bird_count) %>% 
  group_by(species_common_name) %>% 
  summarise(count_of_sightings = n(), 
            total_count = sum(bird_count, na.rm = TRUE)) 
  
# join these onto our original table
seabirds_plus_counts <- left_join(seabirds, 
                                  seabirds_counting,
                                  "species_common_name" )

#filter for individual sightings (count = total count) meaning that each 
# sighting per bird must have been an individual 
task_four_data <- seabirds_plus_counts %>% 
  filter(count_of_sightings == total_count) %>% 
  group_by(species_common_name) %>% 
  summarise(n()) %>% 
  nrow()
task_four_data

# 86 birds only sighted individually
```

### Task 5 How many penguins were seen

```{r}
# Note here I've calculated count of penguins seen not the number of sightings

# here we filter for any names containing penguin then we count the total for the count column
seabirds %>% 
  filter(str_detect(species_common_name,
                    "(?i).*penguin+.*")) %>% 
  summarise(sum(bird_count, na.rm = T))

# 158 penguins were seen
```