---
title: "candy_analysis_doc"
author: "Neil"
date: "2022-07-31"
output: html_document
---
## Load Libraries
```{r}
library(readr)
library(tidyverse)
library(here)
```

## Load Data

```{r}
candy_clean <- read_csv(here("clean_data/candy_clean.csv"))
```

##Tasks

### Task 1 What is the total number of candy ratings given across the three years.

```{r}
# This would be the number of rows in candy_rating != NA
candy_clean %>% 
  drop_na(candy_rating) %>% 
  nrow()

  # total ratings - 767302

```

### Task 2 What was the average age of people who are going out trick or treating?

```{r}

# filter for Yes to going out then average the age column
candy_clean %>% 
  filter(going_out == "Yes") %>% 
  summarise(avg_age = mean(age, na.rm = TRUE))

# Average_age is 35.22
```

### Task 3 What was the average age of people who are not going trick or treating?

```{r}

# filter for No to going out then average the age column
candy_clean %>% 
   filter(going_out == "No") %>% 
summarise(avg_age = mean(age, na.rm = TRUE))

# Average age is 39.26
```

### Task 4 For each of joy, despair and meh, which candy bar received the most of these ratings?

```{r}
# Despair
candy_clean %>% 
  group_by(candy_type) %>% 
  count(candy_rating, name = "count_of_rating") %>% 
  filter(candy_rating == "DESPAIR") %>% 
  ungroup() %>% 
  arrange(desc(count_of_rating)) %>% 
  slice_max(count_of_rating, n = 1)

# Answer : broken_glow_stick

# Joy
candy_clean %>% 
  group_by(candy_type) %>% 
  count(candy_rating, name = "count_of_rating") %>% 
  filter(candy_rating == "JOY") %>% 
  ungroup() %>% 
  arrange(desc(count_of_rating)) %>% 
  slice_max(count_of_rating, n = 1)

# Answer : any_full_sized_candy_bar

# Meh
candy_clean %>% 
  group_by(candy_type) %>% 
  count(candy_rating, name = "count_of_rating") %>% 
  filter(candy_rating == "MEH") %>% 
  ungroup() %>% 
  arrange(desc(count_of_rating)) %>% 
  slice_max(count_of_rating, n = 1)

# Answer : lollipops
```

### Task 5 How many people rated Starburst as despair?

```{r}
# filter for starburst and despair then count the rows
candy_clean %>% 
  filter(candy_type == "starburst" & 
           candy_rating == "DESPAIR") %>% 
  nrow()

# 1990 people thought starburst was despair-worthy
```

#### Change the dataset for new ordering system

```{r}
# add on a column which gives each column a score according to this new scoring method
candy_clean <- candy_clean %>% 
 mutate(candy_rating_new_system = case_when ( candy_rating == "DESPAIR" ~ -1,
                                              candy_rating == "JOY" ~ 1,
                                              TRUE ~ 0
 ))
   
 
```


### Task 6 What was the most popular candy bar by this rating system for each gender in the dataset

```{r}
# Filter for female gender, group by candy type then assign each a value
# according to the sum of the new voting criteria, arrange and slice top hit
candy_clean %>% 
  filter(gender == "Female") %>% 
  group_by(candy_type) %>% 
  summarise(new_rating_total = sum(candy_rating_new_system)) %>% 
  arrange(desc(new_rating_total)) %>% 
  slice_head(n=1)
# Female answer : Any full sized candy bar - 875 count

# Filter for male gender, group by candy type then assign each a value
# according to the sum of the new voting criteria, arrange and slice top hit
candy_clean %>% 
  filter(gender == "Male") %>% 
  group_by(candy_type) %>% 
  summarise(new_rating_total = sum(candy_rating_new_system)) %>% 
  arrange(desc(new_rating_total)) %>% 
  slice_head(n=1)

# Male answer : Any full sized candy bar - 1584 count

```

### Task 7 What was the most popular candy bar in each year?

```{r}
# filter for year, group by candy type, sum the new rating system for each 
# candy type, arrange by highest and slice the top hit
candy_clean %>% 
  filter(year == 2015) %>% 
  group_by(candy_type) %>% 
  summarise(new_rating_total = sum(candy_rating_new_system)) %>% 
  arrange(desc(new_rating_total)) %>% 
  slice_head(n=1)

candy_clean %>% 
  filter(year == 2016) %>% 
  group_by(candy_type) %>% 
  summarise(new_rating_total = sum(candy_rating_new_system)) %>% 
  arrange(desc(new_rating_total)) %>% 
  slice_head(n=1)

candy_clean %>% 
  filter(year == 2017) %>% 
  group_by(candy_type) %>% 
  summarise(new_rating_total = sum(candy_rating_new_system)) %>% 
  arrange(desc(new_rating_total)) %>% 
  slice_head(n=1)

# Answer: Full sized candy bar for each again!
```

### Task 8 What was the most popular candy bar by this rating for people in US, Canada, UK, and all other countries?

```{r}
# filter out the country NA values, group by country and candy type, 
# summarise new rating system by country and candy type
# arrange the total rating for each candy type BY GROUP
# slice the head for each group - return top candy type per country
candy_clean %>% 
  filter(!is.na(country)) %>% 
  group_by(country, candy_type) %>% 
  summarise(new_rating_total = sum(candy_rating_new_system)) %>% 
  arrange(desc(new_rating_total), by_group = TRUE) %>% 
slice(1)
# This returns data where total rating = 1 and are tied.
# I would consider filtering this data for total rating > 1.

candy_clean %>% 
  filter(!is.na(country)) %>% 
  mutate(country = if_else(country %in% c("Usa", "Canada", "United Kingdom"), country, "Other")) %>%   group_by(country, candy_type) %>% 
  summarise(new_rating_total = sum(candy_rating_new_system)) %>% 
  arrange(desc(new_rating_total), by_group = TRUE) %>% 
  slice(1)


```

